---
title: "Iron Supplementation and Vitamin D and IL-6 levels"
author: "Lan Kelly"
engine: knitr
format: 
  docx: default
  html: default
editor: visual
date: "`r Sys.Date()`"
execute: 
  echo: false
bibliography: references.bib
---

# Introduction

This project is an investigation into the association between intravenous iron treatment (IVI) in pregnancy and its effect on post-partum interleukin-6 (IL-6) and Vitamin D levels and hence on maternal depression outcomes HAM-D and EPDS. The data were collected during a clinical trial described in [@froessler2022]. Three forms of Vitamin D were measured: 25(OH)D2, 25(OH)D3 and Epi25(OH)D3.

# Methods

IL-6 and Vitamin D levels were summarised using descriptive statistics. Due to skewness in some of the blood markers, means and standard deviations were calculated from the log-transformed values and then back-transformed. Violin plots of the biomarkers showed the distribution by treatment group and time point (screening, 6 weeks and 12 months post-partum). The biomarkers were affected by lower limits of detection, except for 25(OH)D3.

## Association between IVI and biomarkers

The association between IVI treatment dose (500mg or 1000mg) during pregnancy and biomarkers at 6 weeks and 12 months post-partum was analysed in two ways. The first method used repeated measures linear regression with log-transformed biomarkers as the outcome, with time point by treatment group, stratification group (at randomisation) and log-transformed biomarker value at screening as covariates. Where biomarkers were affected by lower limits of detection, a random number between 0 and the lower limit was generated and used in place of the lower limit. Since interleukin-6 appeared to have two loer limits, 0.9 pg/ml (representing values between 0.5 and 0.9) and 0.2 pg/ml, a random number between 0.5 and 0.9 was used instead of the 0.9 lower limit.

The second method used repeated measures tobit regression, which treats lower or upper limits of detection as a form of censoring. The same covariates as the first method were included in the models. Since there can only be one lower limit, 0.2 was used as the lower limit for IL-6 and 0.9 values were replaced with a random number between 0.5 and 0.9. For 25(OH)D3, there was no lower limit of detection so tobit regression was not used.

Estimates from both models were back-transformed to produced geometric mean rate ratios (RR) and 95% confidence intervals. Results from the two methods were compared for consistency.

## Maternal depression

In the investigation between post-partum maternal depression, biomarkers and IVI treatmetn during pregnancy, biomarkers are treated as a mediator. Some fo the effect of IVI treatment acts directly on maternal depression (the "direct effect"), while the "indirect effect" acts through the biomarkers. However, existing sofware for mediation analysis, such as the R software package 'mediation' [@tingley2014], can only account for lower limits of detection in the **outcome** (maternal depression in this analysis) but not the **mediator** (biomarkers).

\newpage

# Results

## Descriptive Statistics

@tbl-desc-tab shows descriptive statistics for Vitamin D and IL-6 levels at screening, 6 weeks and 12 months post-partum, by randomisation group.

```{r}
#| echo: false
#| include: false
#| warning: false
source("vitd_il6_regression.R")
```

```{r}
#| label: tbl-desc-tab
#| tbl-cap: Descriptive Statistics
#| results: 'asis'

desc_wide %>%
  select(FU, il_6_BLUE,il_6_PINK,epi25ohd3_BLUE, epi25ohd3_PINK,
         x25ohd2_BLUE, x25ohd2_PINK, x25ohd3_BLUE, x25ohd3_PINK) %>%
  mutate(FU = case_when(FU == "SCR" ~ "Screening",
                        FU == "FU1" ~ "6 weeks PP",
                        FU == "FU2" ~ "12 months PP")) %>%
  flextable() %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(FU = "Time point",
                    il_6_BLUE = "IL-6, \n500 mg",
                    il_6_PINK = "IL-6, \n1000 mg",
                    epi25ohd3_BLUE = "Epi-25(OH)D3, \n500mg",
                    epi25ohd3_PINK = "Epi-25(OH)D3, \n1000mg",
                    x25ohd2_BLUE = "25(OH)D2, \n500mg",
                    x25ohd2_PINK = "25(OH)D2, \n1000mg",
                    x25ohd3_BLUE = "25(OH)D3, \n500mg",
                    x25ohd3_PINK = "25(OH)D3, \n1000mg"
                    ) %>%
  set_caption(caption = "Biomarkers (mean (SD))")




```

\newpage

## Figures

Violin plots are shown of the distribution of values for IL-6 and three forms of Vitamin D. All show the effect of lower limits of detection except 25(OH)D3.

### Interleukin-6

@fig-il6 shows IL-6 by time point. SCR is screening, FU1 and FU2 are 6 weeks and 12 months post-partum.

```{r}
#| label: fig-il6
#| fig-cap: IL-6 by time point and treatment group
#| warning: false



plotdat <- bind_rows(base %>% mutate(FU="SCR"), 
                     mh_il6_vitd %>%
                       rename(x25ohd3=x25ohd3fu,
                              epi25ohd3=epi25ohd3fu,
                              x25ohd2=x25ohd2fu)) %>%
  mutate(rand = ifelse(rand=="BLUE", "500mg", "1000mg"),
         rand=relevel(factor(rand), ref="500mg"),
         FU = relevel(factor(FU), ref="SCR"))
  
  plotdat %>%
  ggplot(aes(x=rand, y=il_6)) +
    geom_violin() +
    facet_wrap(~FU) +
    ylab("IL-6 (pg/ml)") + xlab("") +
    theme_bw()


```

\newpage

### Vitamin D

@fig-d2, @fig-d3 and @fig-epid3 are violin plots of the three different forms of Vitamin D, 25(OH)D2, 25(OH)D3 and Epi25(OH)D3.

```{r}
#| label: fig-d2
#| fig-cap: 25(OH)D2 by time point and treatment group
#| warning: false


plotdat %>%
  ggplot(aes(x=rand, y=x25ohd2)) +
    geom_violin() +
    facet_wrap(~FU) +
    ylab("25(OH)D2") + xlab("") +
    theme_bw()



```

```{r}
#| label: fig-d3
#| fig-cap: 25(OH)D3 by time point and treatment group
#| warning: false


plotdat %>%
  ggplot(aes(x=rand, y=x25ohd3)) +
    geom_violin() +
    facet_wrap(~FU) +
    ylab("25(OH)D3") + xlab("") +
    theme_bw()
```

```{r}
#| label: fig-epid3
#| fig-cap: Epi25(OH)D3 by time point and treatment group
#| warning: false


plotdat %>%
  ggplot(aes(x=rand, y=x25ohd3)) +
    geom_violin() +
    facet_wrap(~FU) +
    ylab("Epi25(OH)D3") + xlab("") +
    theme_bw()
```

\newpage

## Regression

@tbl-tobit shows the results of regression for tobit vs random methods. "Censored" refers to the lower limit of detection analysed using the tobit model.

Results from the two method were very similar. There were no significant differences in biomarkers between the two treatment groups at any time point, although the RR for IL-6 levels at 6 weeks was furthest from 1.0 and may have been significant if the sample size was larger.

```{r}
#| label: tbl-tobit
#| results: 'asis'
#| tbl-cap: Regression output


tab_cens %>%
  select(Outcome, Contrast, Level, est_ci, p) %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(est_ci = "RR (95% CI)",
                    p="P-value")



```

\newpage

# Mediation analysis

Mediation analysis aimed to estimate the differences between two doses of IVI on maternal mental health outcomes (HAM-D and EPDS) ("total effect") through two mechanisms:

1.  the direct effect of IVI treatment, or average direct effect (ADE)

2.  the indirect effect of IVI treatment through the mediation of biomarkers IL-6 or 25(OH)D3, or the average causal mediation effect (ACME)

The random number method replaced the biomarker values affected by lower limits of detection for IL-6. For Vitamin D, the mediation analysis used only 25 (OH)D3 and not the other two forms.

```{r}
#| echo: false
#| include: false
#| warning: false

source("VitD_IL6_mh_mediate.R")

```

\newpage

## Descriptive Statistics

Descriptive Statistics for 25(OH)D3, HMAD and EPDS at screening, 6 weeks and 12 months. **The numbers included in the descriptive statistics are different to @tbl-desc-tab because participants with missing data in either the biomarker or the outcome have been excluded.** There is more missing data for EPDS than for HAM-D. The number of participants for EPDS at 12 months at particularly low.

### 25(OH)D3

#### HAM-D

```{r}
#| label: tbl-hamd-25ohd3-desc
#| tbl-cap: "Mean (SD) for 25(OH)D3 and HAM-D at screening (baseline), 6 weeks and 12 months. HAMD at 6 weeks and 12 months is change from baseline."

hamd_dep_25ohd3_fu1$desc %>%
  dplyr::select(timept, x25ohd3fu_1, x25ohd3fu_0, 
                mh_hamd_dep_1, mh_hamd_dep_0) %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(timept = "Time point",
                    x25ohd3fu_0="25(OH)D3, \n1000mg",
                    x25ohd3fu_1="25(OH)D3, \n500mg",
                    mh_hamd_dep_0="HAM-D, \n1000mg",
                    mh_hamd_dep_1="HAM-D, \n500mg"
                    ) %>%
  footnote(i=c(2,3,2,3), j=c(4,4,5,5),
           value = as_paragraph("Change from baseline"),
  ref_symbols = "*",
  part = "body",
  inline = T,
  sep = "")

```

#### EPDS

```{r}
#| label: tbl-epds-25ohd3-desc
#| tbl-cap: "Mean (SD) for 25(OH)D3 and EPDS at screening (baseline), 6 weeks and 12 months. EPDS at 6 weeks and 12 months is change from baseline."

epds_25ohd3_fu1$desc %>%
  dplyr::select(timept, x25ohd3fu_1, x25ohd3fu_0, 
                mh_epds_1, mh_epds_0) %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(timept = "Time point",
                    x25ohd3fu_0="25(OH)D3, \n1000mg",
                    x25ohd3fu_1="25(OH)D3, \n500mg",
                    mh_epds_0="EPDS, \n1000mg",
                    mh_epds_1="EPDS, \n500mg"
                    ) %>%
  footnote(i=c(2,3,2,3), j=c(4,4,5,5),
           value = as_paragraph("Change from baseline"),
  ref_symbols = "*",
  part = "body",
  inline = T,
  sep = "")
```

\newpage

### IL-6

#### HAM-D

```{r}
#| label: tbl-hamd-il6-desc
#| tbl-cap: "Mean (SD) for IL-6 and HAM-D at screening (baseline), 6 weeks and 12 months. HAMD at 6 weeks and 12 months is change from baseline."

hamd_dep_il6_fu1$desc %>%
  dplyr::select(timept, log_il6_1, log_il6_0, 
                mh_hamd_dep_1, mh_hamd_dep_0) %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(timept = "Time point",
                    log_il6_0="IL-6, \n1000mg",
                    log_il6_1="IL-6, \n500mg",
                    mh_hamd_dep_0="HAM-D, \n1000mg",
                    mh_hamd_dep_1="HAM-D, \n500mg"
                    ) %>%
  footnote(i=c(2,3,2,3), j=c(4,4,5,5),
           value = as_paragraph("Change from baseline"),
  ref_symbols = "*",
  part = "body",
  inline = T,
  sep = "")



```

#### EPDS

```{r}
#| label: tbl-epds-il6-desc
#| tbl-cap: "Mean (SD) for IL-6 and EPDS at screening (baseline), 6 weeks and 12 months. EPDS at 6 weeks and 12 months is change from baseline."

epds_il6_fu1$desc %>%
  dplyr::select(timept, log_il6_1, log_il6_0, 
                mh_epds_1, mh_epds_0) %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(timept = "Time point",
                    log_il6_0="IL-6, \n1000mg",
                    log_il6_1="IL-6, \n500mg",
                    mh_epds_0="EPDS, \n1000mg",
                    mh_epds_1="EPDS, \n500mg"
                    ) %>%
  footnote(i=c(2,3,2,3), j=c(4,4,5,5),
           value = as_paragraph("Change from baseline"),
  ref_symbols = "*",
  part = "body",
  inline = T,
  sep = "")

```

\newpage

## Mediation Results

Estimates for ACME, ADE and Total Effect are the difference in the outcome for 500mg-1000mg. Maternal mental health outcomes are the change in HAM-D or EPDS from baseline.

```{r}
#| label: tbl-med-results

tab_med = bind_rows(df_hamd_dep_25ohd3, df_epds_25ohd3,
                    df_hamd_dep_il6, df_epds_il6)

tab_med = tab_med %>%
  dplyr::select(Outcome, Mediator, timept, acme, acme_p,
                ade, ade_p, ate, ate_p,
                prop_med)

tab_med %>%
  flextable %>%
  autofit() %>%
  theme_box() %>%
  set_header_labels(timept = "Time point",
                    acme = "ACME (95% CI)", 
                    acme_p = "ACME, p-value",
                    ade = "ADE (95% CI)", 
                    ade_p = "ADE, p-value", 
                    ate = "Total effect (95% CI)", 
                    ate_p = "Total effect, p-value",
                    prop_med = "Proportion mediated") 

```

\newpage

## Plots

### 25(OH)D3

#### HAM-D

```{r}
#| label: fig-25ohd3-hamd
#| fig-height: 7
#| fig-dpi: 300
#| fig-cap: "Average Causal Mediation Effect (ACME), Average Direct Effect (ADE) and Total Effect for HAM-D mediated by 25(OH)D3"

par(mfrow = c(2,1))

plot(hamd_dep_25ohd3_fu1$medfit,  main = "6 weeks",xlim = c(-5,1),
    treatment = "treated",
     # labels  =c("Indirect","Direct" ,"Total"),
     xlab = "Change in HAM-D from baseline (500mg-1000mg)")

plot(hamd_dep_25ohd3_fu2$medfit,  main = "12 months",xlim = c(-5,1),
    treatment = "treated",
     #labels  =c("Indirect","Direct" ,"Total"), 
     xlab = "Change in HAM-D from baseline (500mg-1000mg)")


```

\newpage

#### EPDS

```{r}
#| label: fig-25ohd3-epds
#| fig-height: 7
#| fig-dpi: 300
#| fig-cap: "Average Causal Mediation Effect (ACME), Average Direct Effect (ADE) and Total Effect for EPDS mediated by 25(OH)D3"


par(mfrow = c(2,1))

plot(epds_25ohd3_fu1$medfit,  main = "6 weeks",xlim = c(-4,6),
    treatment = "treated",
     # labels  =c("Indirect","Direct" ,"Total"),
     xlab = "Change in EPDS from baseline (500mg-1000mg)")

plot(epds_25ohd3_fu2$medfit,  main = "12 months",xlim = c(-4,6),
    treatment = "treated",
     #labels  =c("Indirect","Direct" ,"Total"), 
     xlab = "Change in EPDS from baseline (500mg-1000mg)")

```

\newpage

### IL-6

#### HAM-D

```{r}
#| label: fig-il6-hamd
#| fig-height: 7
#| fig-dpi: 300
#| fig-cap: "Average Causal Mediation Effect (ACME), Average Direct Effect (ADE) and Total Effect for HAM-D mediated by IL-6"

par(mfrow = c(2,1))

plot(hamd_dep_il6_fu1$medfit,  main = "6 weeks",xlim = c(-5,1),
    treatment = "treated",
     # labels  =c("Indirect","Direct" ,"Total"),
     xlab = "Change in HAM-D from baseline (500mg-1000mg)")

plot(hamd_dep_il6_fu2$medfit,  main = "12 months",xlim = c(-5,1),
    treatment = "treated",
     #labels  =c("Indirect","Direct" ,"Total"), 
     xlab = "Change in HAM-D from baseline (500mg-1000mg)")



```

\newpage

#### EPDS

```{r}
#| label: fig-il6-epds
#| fig-height: 7
#| fig-dpi: 300
#| fig-cap: "Average Causal Mediation Effect (ACME), Average Direct Effect (ADE) and Total Effect for EPDS mediated by IL-6"

par(mfrow = c(2,1))

plot(epds_il6_fu1$medfit,  main = "6 weeks",xlim = c(-4,6),
     treatment = "treated",
     #labels  =c("Indirect","Direct" ,"Total"),
     xlab = "Change in EPDS from baseline (500mg-1000mg)")

plot(epds_il6_fu2$medfit,  main = "12 months",xlim = c(-4,6),
    treatment = "treated",
     #labels  =c("Indirect","Direct" ,"Total"), 
     xlab = "Change in EPDS from baseline (500mg-1000mg)")

par(mfrow = c(1,1))
```

# Discussion

There was a significant difference in the total effect and average direct effect (ADE) between the 500mg and 100mg groups on HAM-D and EPDS at 6 weeks. **However, there was no significant average causal mediation effect (ACME) from 25(OH)D3 or IL-6 on either maternal mental health outcome. In addition, the proportion of the total effect that was mediated was close to zero for all mediators and outcomes.**

The results for HAMD-D and particularly EPDS at 12 months are variable depending on the mediator, most likely due to different patterns of missing data between IL-6 and 25(OH)D3. **Therefore, results at 12 months may be biased due to missing data and should be treated with caution.**

# References
